Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-04-27 manuscript.Rmd
Lots of references fixed

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [$CO_2$] but tracks water availability
author: "Remko Duursma"
date: "April 24, 2015"
output:
  word_document:
    reference_docx: msstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

@@

Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency (Ainsworth and Long, 2005; Ellsworth et al., 2012). Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) (Norby and Zak, 2011). Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* (Norby and Zak, 2011), although positive responses to $eC_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak (2011) found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$. Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger (Norby et al., 2005)

In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* (Donohue et al., 2009), since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area (Warren et al., 2010) leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.
In support of this prediction, (Donohue et al., 2013) attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$. Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.

Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g. (Richardson et al., 2012; Savoy and Mackay, 2015) whereas *L* responses to fluctuations in water availability are qualitatively well-known (more cites), but poorly quantified. The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability (Pook, 1985; Specht and Specht, 1989; Whitehead and Beadle, 2004)(Pook et al., 1997)(Limousin et al., 2009).The strong relationship between *L* and rainfall observed across Australia (Ellis and Hatton, 2008; Specht and Specht, 1989), and other continents (Donohue et al., 2013; Grier and Running, 1977; [@Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales (Pook, 1985; Pook et al., 1997), driven by leaf turnover (Misson et al., 2011) and leaf flushing and expansion (Muller et al., 2011), which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology (Misson et al., 2011).

We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$. In addition, we expected that the low *L* that is typical of native Eucalyptus woodlands (Whitehead and Beadle, 2004) allowed a larger response to $eC_a$ (Norby and Zak, 2011). Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

We then estimated leaf area index (*L*) from τ_d (*L*τ), following (Campbell and Norman, 2000). We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where θ is the solar zenith angle, G(θ) is the foliage projection function in the direction of the beam (Campbell and Norman, 2000). Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 (Stenberg, 2006). We solved numerically for *L* in Eq. 3, using a smooth integration over θ, yielding *L*τ.

To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate τ_d  were available during this period. In theory, cumulative *L* leaf litter production (PL, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated equation (4),

@@

where *c* is a calibration parameter, Lt=0 is $L_\tau$ at the start of the drought period. We used linear regression without intercept to estimate c in Eq. 4 as `r calib` (SE = `r tidy(caliblm)$std.error`). It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq. 4.

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014). For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.

@@

An outbreak of psyllids (Cardiaspina sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in Eucalyptus moluccana, see (Hall et al., 2015)). This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower τ_d than before the insect attack.

@@

## Response of *L* to elevated [$CO_2$]
We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen Eucalyptus woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$. We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011). Counter to our expectations but similar to Norby et al. (2003), we found no significant effect of elevated $C_a$ on *L*. It could be argued that the change in *L* in response to $eC_a$ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a 6% in *L*. Based on this small detectable difference, we conclude that the lack of a significant $C_a$ effect is not due to a lack of power, but a true lack of response to elevated $C_a$.

This lack of a response to elevated $C_a$ counters our expectations and those in previous publications (Norby et al. 2003 McCarthy et al. 2010 etc.), for a number of reasons. First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to $eC_a$. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. 1.4 m$^2$ m$^{-2}$, which is low compared to a global compendium of forest *L* (Iio et al., 2013), and the lowest *L* of any forest FACE experiment (Norby and Zak, 2011), but in line with a range of values for native Eucalyptus forests and woodlands reported by (Whitehead and Beadle, 2004) and (Iio et al., 2013). Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under $eC_a$. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands (White et al., 2010).

@@

## Litter production and dynamics of *L*
We found a correlation between *L* dynamics and litter production, both during periods of decreasing and increasing *L*. During a sustained period without significant rainfall (42 mm over 4 months), we used the decrease in *L* as an opportunity to calibrate against litter production. During periods of increasing *L*, litter fall increased sharply, indicating replacement of old leaves by young leaves. Indeed we observed in the field that after the significant flushing event of Dec. 2013, very few old leaves were present in the canopy. This is in line with (Pook, 1984a) in Corymbia maculata (a genus closely related to Eucalyptus), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months. Similarly, (Stone and Bacon, 1995) showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale. Since leaf flushing has been shown to be closely related to soil water availability (Fig. 7; Pook, 1984a), net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing. The consequence is that leaf turnover is highly variable in time.

## Consequences for phenology models
Current-generation phenology models are heavily biased towards temperate conditions (Primack and Miller-Rushing, 2011), where leaf flushing occurs in spring in response to increase in temperature (Richardson et al., 2012; Savoy and Mackay, 2015). In contrast, phenology of evergreen Eucalyptus may be more opportunistic, and shows rapid response to water availability (Fig. 7). (Specht and Brouwer, 1975) demonstrated for four Eucalyptus species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth. Similarly, (Pook, 1984b) demonstrated leaf production during most months in Corymbia maculata, except during winter and suppression by dry periods (Pook, 1984a). Thus a minimum temperature is also required for leaf growth to occur. We also have not observed leaf flushing in winter time (Ellsworth et al., unpubl. data), but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig. 7).

>>>>

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [$CO_2$] but tracks water availability
author: "Remko Duursma"
date: "April 24, 2015"
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

@@

Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* [@Norby2011], although positive responses to $eC_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$. Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger [@Norby2005].

In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* [@Donohue2009], since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. **Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$. Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.

Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g. [@Richardson2012; @Savoy2015]) whereas *L* responses to fluctuations in water availability are qualitatively well-known (**more cites**), but poorly quantified. The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009].The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].

We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to $eC_a$ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

We then estimated leaf area index (*L*) from $\tau_d$ ($L_\tau$), following (Campbell and Norman, 2000). We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam (Campbell and Norman, 2000). Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 (Stenberg, 2006). We solved numerically for *L* in Eq. 3, using a smooth integration over $\theta$, yielding *L*τ.

To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. In theory, cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated equation (4),

@@

where *c* is a calibration parameter, $L_{\tau,t=0}$ is $L_\tau$ at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r calib` (SE = `r tidy(caliblm)$std.error`). It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq. 4.

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries (Wood, 2006), using the *mgcv* package in R [@RCT2015]. For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.

@@

An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see (Hall et al., 2015)). This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower τ_d than before the insect attack.

@@

## Response of *L* to elevated [$CO_2$]
We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$. We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011). Counter to our expectations but similar to Norby et al. (2003), we found no significant effect of elevated $C_a$ on *L*. It could be argued that the change in *L* in response to $eC_a$ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a 6% in *L*. Based on this small detectable difference, we conclude that the lack of a significant $C_a$ effect is not due to a lack of power, but a true lack of response to elevated $C_a$.

This lack of a response to elevated $C_a$ counters our expectations and those in previous publications (Norby et al. 2003 McCarthy et al. 2007), for a number of reasons. First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to $eC_a$. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. 1.4 m$^2$ m$^{-2}$, which is low compared to a global compendium of forest *L* (Iio et al., 2013), and the lowest *L* of any forest FACE experiment (Norby and Zak, 2011), but in line with a range of values for native Eucalyptus forests and woodlands reported by (Whitehead and Beadle, 2004) and (Iio et al., 2013). Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under $eC_a$. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands (White et al., 2010).

@@

## Litter production and dynamics of *L*
We found a correlation between *L* dynamics and litter production, both during periods of decreasing and increasing *L*. During a sustained period without significant rainfall (42 mm over 4 months), we used the decrease in *L* as an opportunity to calibrate against litter production. During periods of increasing *L*, litter fall increased sharply, indicating replacement of old leaves by young leaves. Indeed we observed in the field that after the significant flushing event of Dec. 2013, very few old leaves were present in the canopy. This is in line with (Pook, 1984a) in *Corymbia maculata* (a genus closely related to *Eucalyptus*), who showed accelerated litter production during periods of net *L* increase, but with a lag between litter fall and production of about 2 months. Similarly, (Stone and Bacon, 1995) showed that drought-stressed trees had higher litter production at the branch scale, but irrigated trees had larger litter production at the canopy scale. Since leaf flushing has been shown to be closely related to soil water availability (Fig. 7; Pook, 1984a), net leaf turnover is driven by water availability both through increased litter production during extended droughts, and complete canopy replacement following leaf flushing. The consequence is that leaf turnover is highly variable in time.

## Consequences for phenology models
Current-generation phenology models are heavily biased towards temperate conditions (Primack and Miller-Rushing, 2011), where leaf flushing occurs in spring in response to increase in temperature (Richardson et al., 2012; Savoy and Mackay, 2015). In contrast, phenology of evergreen Eucalyptus may be more opportunistic, and shows rapid response to water availability (Fig. 7). (Specht and Brouwer, 1975) demonstrated for four Eucalyptus species that leaf growth  peaked both in spring and autumn, with low winter temperatures and high summer temperatures (and low water availability) suppressing leaf growth. Similarly, (Pook, 1984b) demonstrated leaf production during most months in *Corymbia maculata*, except during winter and suppression by dry periods (Pook, 1984a). Thus a minimum temperature is also required for leaf growth to occur. We also have not observed leaf flushing in winter time (Ellsworth et al., unpubl. data), but at our site this was confounded with water availability as both winters during the study period were relatively dry (Fig. 7).
